---
title: "course challenge"
author: "Awesome Me"
date: "2023-04-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load libraries


```{r }
library(tidyverse)
library(here)
```

## import flavors_of_cacao.csv file


```{r}
chocolate_df <- read_csv(here("data", "flavors_of_cacao.csv"))
chocolate_df
```

## review the column names

```{r}
colnames(chocolate_df)
```

## use the select() function to create a new data frame with only these three variables. 
### (Rating, CocoaPercent, BeanType)

```{r}
trimmed_flavors_df <- chocolate_df %>% select(Rating, CocoaPercent, BeanType)
trimmed_flavors_df
```

## use the summarize() and sd() functions to find the standard deviation of the rating for your data. 
### Add the code chunk that lets you find the standard deviation for the variable Rating.

```{r}
trimmed_flavors_df <- chocolate_df %>% summarise(sd(Rating))
trimmed_flavors_df
```

## apply the filter() function to the variables CocoaPercent and Rating
### filter the data frame for chocolate bars that contain at least 75% cocoa and have a rating of at least 3.9 points. 

```{r}
best_trimmed_flavors_df <- chocolate_df %>% filter(CocoaPercent >= 0.75, Rating >= 3.9)
best_trimmed_flavors_df
```

## TODO: Question 8




